\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data}}

### 2012-03-05 package/smutans/inst/doc/smutans.Rnw
Just an update.

\begin{abstract}
We analyze RNA samples from \textit{Streptococcus mutans} using
RNA-seq to define genes whose expression changes between samples
\cite{Oshlack2010}. Short reads are first mapped to a reference genome using
approaches such as \program{BWA} \cite{Li2009a} or \program{subread}.
Reads fitting into one of the annotated
intervals are counted.  These counts are passed to a statistical package which
detects differences in read counts for statistical significance. For this
purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
\cite{Robinson2010}. This prodecure returns a list of differentially expressed
genes, which can be passed to downstream analysis, such as functional category
enrichment test using \Rpackage{goseq} \cite{Young2010}.  The current
\Rpackage{smutans} contains procedures that use functions provided by
the above R packages. 
\end{abstract}

@@

Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would not be mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We trimmed sequence by quality score from the 3' end
with threshold score of 20 \cite{Schmieder2011}.
We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.  To obtain count tables of short reads we
used scripts available at 
\textit{\http://github.com/goshng/RNASeq-Analysis}. In the following we will use
the output of short reads

The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
which is a type supported by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
\program{samtools} \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
annotations in the UA159 genome we counted short reads that overlapped the
genes.  The count values are raw counts of sequencing reads. The
\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
reads that allows to assess the measurement precision correctly; we did not have
rounded values of normalized counts, or counts of covered base pairs.  Several
exeriments of \textit{S.\ mutans} RNA sampling were performed.  \fixme{Find the
papers that are related with the experiments.} One of them was presented in the
\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the data type used
by \Rpackage{DESeq}. We have a convenient \Rclass{Smutans}, which contains slots
for \Rclass{CountDataSet} objects, and results from analyses thereof.  Before
loading necessary R packages and data sets, let us describe what functions were
used for preparing data objects such as \Robject{smutansGenes}. One should
create an R script file at the directory that contains the source code directory
of the \Rpackage{smutans}. Put the following scripts in the R script file and
run the script at that directory. That should create RData files at the
directory smutans/data. User should not run the R script because the data files
should be found in the \Rpackage{smutans}.

% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
% \fixme{Where are flat files for genes of S. mutans UA159?}
<<samplePreparation,eval=FALSE>>=
# not run
# but create a R scirpt file with the following scripts, and
# run it, i.e., Rscript run.R
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData34( )
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
smutans.prepareDataOMZ175( ) 
smutans.prepareDataUA159CSP( ) 
smutans.prepareDataSMU86CSP( )
smutans.prepareTranscript( )
q("no")
@
<<options,results=hide,echo=FALSE>>=
options(digits=3)
@
Now, see some of the data objects by loading the \Rpackage{smutans}, and a data
\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
\Rpackage{DESeq} in which users should find \Rfunction{counts} and
\Rfunction{pData}. 
<<smutans,results=hide>>=
library( "smutans" )
data( "smutansGenes" )
@
There are 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA-seq samples in \Robject{smutansGenes}, which
contains the counts and also metadata about the samples:
<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
@
<<look_into_smutansGenes>>=
pData( smutansGenes )
@
The samples differ by experimental condition (``glucose'' or
``galactose'') and by library type (``ua159'' or ``tw1'').

Samples of 835NP, 835P, and UA159 no CSP should be loaded using the following
command.
<<sm835,results=hide>>=
data( "sm835Genes" )
@

% Resources
% Google charts - Venn Diagram
% http://code.google.com/apis/chart/index.html

\fixme{Describe other data objects as well.
sm34Genes.RData,
sm835Genes.RData,
smutans.cat.desc.RData,
smutans.feature.genes.RData, and
smutans.go.genes.RData.}

%--------------------------------------------------
\section{DE analysis using UA159 and TW1 samples with
glucose and galactose} \label{sec:detw1}
%--------------------------------------------------

We could compare four pairwise comparisons using the two-factor data set of
\Robject{smutansGenes}: a pair of comparisons between ``glucose'' and
``galactose'' given either type ``ua159'' or ``tw1'', and another pair of
comparisons between types ``ua159'' and ``tw1'' given either condition
``glucose'' or ``galactose.'' We focus on one of the four pairwise comparisons
first. Later, we will perform analyses for the rest of the three comparisons.
We look at the ``ua159'' data for now, and will look at the ``tw1'' data later.

We can cluster samples using variance stabilized data. Figure \ref{figClust}
shows the \Sexpr{length(colnames(counts(smutansGenes)))} sampled clustered. 
The clustering should reflect our experimental design, i.e., samples are
more similar when they have the same treatment or the same library type.
To make this conclusion, we estimate the dispersion with
mode ``blind'' (see \Rpackage{DESeq} vignette), as only then,
the variance stabilizing transformation is not informed about the design and
hence not biased towards a result supporting it. Such an analysis is useful
for quality control.  This appears to be useful as a first in any analysis.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilizing transformation of the
  count data.}
\label{figClust}
\end{figure}
<<figClust,fig=TRUE>>=
allSample <- newSmutans( smutansGenes )
smutans.de2Clust( allSample )
@
As expected, two experimental conditions of ``glucose'' and ``galactose'' divide the
samples. Within each condition, two different types of ``ua159'' and ``tw1'' divide
them again. The pair of ``glucose ua159'' and ``galactose tw1'' is the most
distinguished.

%------------------------------------------------------------
\subsection{Inference using single-factor data} 
\label{subsec:ua159only}
%------------------------------------------------------------

We create a \Rclass{Smutans} object using the \Rclass{CountDataSet}
\Robject{smutansGenes}. The title option would be useful in distinguishing
different objects of \Rclass{Smutans} as we have more R objects with the type.  
A subsequent call to
\Rfunction{smutans.de2} uses a two-factor data set to test differential
expression of genes.  The \Rfunction{smutans.de2} estimates effective library
sizes, and dispersions using functions provided by \Rpackage{DESeq}. Then, it
tests genes for differential expression.
<<ua159onlySamples>>=
ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
ua159only <- smutans.de2( ua159only, type="ua159", 
                          condA="glucose", condB="galactose" )
@
Because \Rpackage{DESeq} recommended to check
the dispersion estimates and to make sure that the data quality is as
expected, we plot the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve in red color
(Fig.\ \ref{figFit}).

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFit}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength for the data set of UA159-only sample
to compare glucose and galactose conditions.}
\label{figFit}
\end{figure}
<<figFit,fig=TRUE>>=
smutans.plotDispersionEstimates(ua159only)
@
\fixme{The following paragraph is adopted from \Rpackage{DESeq} vignette.}
As we estimated the dispersion from only two samples, we should expect the
estimates to scatter with quite some sampling variance around their true values.
Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
test, because using too low dispersion values leads to false positives.  Many of
the values below the red line are likely to be underestimates of the true
dispersions, and hence, it is prudent to instead rather use the fitted value.
On the othe hand, not all of the values above the red line are overestimations,
and hence, the conservative choice is to keep them instead of replacing them
with their fitted values.
Having estimated the dispersion for each gene, we look for differentially
expressed genes. To contrast two conditions, e.g., to see whether there is
differential expression between conditions ``glucose'' and ``galactose'',
\Rfunction{smutans.de2} already performed the 
\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  
Let's list genes with resulting information from \Rpackage{DESeq}.
<<ua159onlyDE2List>>=
head(smutans.de2List( ua159only ))
@
For each gene, we get its mean expression level (at the base scale) as a joint
estimate from both conditions, and estimated separately for each condition, the
fold change from the first to the second condition, i.e., 
$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2)
of the fold change, and the $p$ value for the statistical significance of this
change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
testing with the Benjamini-Hochberg procedure (see the R function
\Rfunction{p.adjust}), which controls false discovery rate (FDR). Later, we
print the gene list out in a CSV file using option file of
\Rfunction{smutans.de2List}.
%-----------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figDE}
\caption{Plot of normalised mean versus $log_2$ fold change
(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
versus ``galactose'' using UA159-only sample.}
\label{figDE}
\end{figure}
%-----------------------------------------------------------------------------
<<figDE,fig=TRUE>>=
smutans.plotDiffExp(ua159only)
@
Figure \ref{figDE} shows a plot of the $\log_2$ fold changes against the base
means, coloring in red those genes that are significant at 10\% of FDR. 
Let us use ``tw1'' type to compare ``glucose'' and ``galactose'' conditions.
We continue to analyze the data set using type ``tw1'' to compare the two
conditions. We perform the single-factor analysis. 
<<tw1onlySamples>>=
tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
tw1only <- smutans.de2( tw1only, type="tw1", 
                        condA="glucose", condB="galactose" )
tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
addmargins( tabUA159onlyTW1only )
@
The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
genes were found in both analyses. The analysis using TW1-only sample found
\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using UA159-only sample found
\Sexpr{tabUA159onlyTW1only[2,1]} genes.

Until now we focused on the effect of condition either using UA159-only or
TW1-only samples. We now change the gear to see effects of sample types on
differentially expressed genes. We create two \Rclass{Smutans} objects:
\Robject{glucoseOnly} and \Robject{galactoseOnly}, and compare them in detecting
differentially expressed genes:
<<glucoseOnlySamples>>=
glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
                              condA="ua159", condB="tw1" )
tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
                                                      galactoseOnly )
addmargins( tabGlucoseOnlyGalactoseOnly )
@
Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
rate of 10\%. The analysis using galactose-only sample found
\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose-only
sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes. 
Table \ref{tab:de} summarizes the number of differentially expressed genes that
were found using single-factor analyses at FDR of 10\%.

%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Numbers of differentially expressed genes in different analyses}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{ccccl}
\hline 
Sample A & Sample B & Number of Genes & Result File \\
\hline 
UA159 Glucose & UA159 Galactose & \Sexpr{sum(ua159only@padj < .1,na.rm=T)} & ua159only.csv \\
TW1 Glucose & TW1 Galactose & \Sexpr{sum(tw1only@padj < .1,na.rm=T)} & tw1only.csv \\
UA159 Glucose & TW1 Glucose & \Sexpr{sum(glucoseOnly@padj < .1,na.rm=T)} & glucoseOnly.csv \\
UA159 Galactose & TW1 Galactose & \Sexpr{sum(galactoseOnly@padj < .1,na.rm=T)} & galactoseOnly.csv \\
% UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
% TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
% UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
% UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:de}
\end{table}
%-----------------------------------------------------------------------------

%------------------------------------------------------------
\subsection{Inference using all of the samples} 
\label{subsec:allSample}
%------------------------------------------------------------

>>>>

\begin{abstract}
We analyze RNA samples from \textit{Streptococcus mutans} using
RNA-seq to define genes whose expression changes between samples
\cite{Oshlack2010}. Short reads are first mapped to a reference genome using
approaches such as \program{BWA} \cite{Li2009a} or \program{subread}.
Reads fitting into one of the annotated
intervals are counted.  These counts are passed to a statistical package which
detects differences in read counts for statistical significance. For this
purpose, we make use of \Rpackage{DEseq} \cite{Anders2010}.
This prodecure returns a list of differentially expressed
genes, which can be passed to downstream analysis, such as functional category
enrichment test using \Rpackage{goseq} \cite{Young2010}.  The current
\Rpackage{smutans} contains procedures that use functions provided by
the above R packages. 
\end{abstract}

@@

Analyzing RNA-seq data for determining differentially expressed genes
are well described \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference
genome, \textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because
the aligner allowed a few gaps for efficient alignment of millions of sequences
of size being 100 base pairs long, short reads that consist of RNAs and
sequencing adapter sequences would not be mapped. By using \program{cutadapt}
\cite{Martin2011} we removed parts of adapter sequences from short reads before
mapping on the reference genome.  We also trimmed short read sequences by quality score from
the 3' end with threshold score of 20 \cite{Schmieder2011}. Mapped short read
alignments were converted into readable formats \cite{Li2009b}. We counted short
reads aligned to annotated genes in the reference genome, forming a table of
read counts of the genes. Statistical software R \cite{R2011} with
\Rpackage{DESeq} \cite{Anders2010} was subsequently employed to infer
differentially expressed in different biological conditions. Most of functions
in comparing different experiments were derived from \Rpackage{DESeq}. Readers
should refer to the package for details.  We also generated
tracks for a Streptococcus Genome Browser
(http://strep-genome.bscb.cornell.edu) that summarize gene expression levels
based on the short read alignments. This allowed us to visualize parts of the
reference genome for comparing different RNA samples. Further analyses such as
functional category enrichment test \cite{Young2010} were also employed.  To
obtain count tables of short reads we used scripts available at a website
http://github.com/goshng/RNASeq-Analysis. In the following we will use
the count table to determine differentially expressed genes.

The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
which is a type supported by \Rpackage{DESeq}.  Note that the count values
are raw counts of sequencing reads.  The \Rpackage{DESeq}'s statistical model,
or negative binomial model, holds for raw reads counts that allows to assess the
measurement precision correctly; we did not have rounded values of normalized
counts, or counts of covered base pairs to form count tables.  Several
exeriments of \textit{S.\ mutans} 
RNA sampling were performed to examine effects of glucose and galactose,
835NP and 835P, OMZ175 and OMZ/HKRR, CSP, and acid tolerance.  The comparison of
glucose and galactose is presented in the \Robject{smutansGenes} as a part of
\Rpackage{smutans}.

<<options,results=hide,echo=FALSE>>=
options(digits=3)
@
In the following chapter, we use \Robject{smutansGenes}, which is of type
\Rclass{CountDataSet} of \Rpackage{DESeq}.  Load the \Rpackage{smutans}, and the
data \Robject{smutansGenes}. 
<<smutans,results=hide>>=
library( "smutans" )
@

Browse the installed directory of \Rpackage{smutans} for several count
table and index files. As we go along the chapters of this vignette, we will use
them.
<<readSmutans>>=
inDir <- system.file("extdata", package="smutans", mustWork=TRUE)
dir(inDir,pattern="count*")
@

Look at two kinds of files: count.txt and count.txt.index. We use
\Rfunction{readSmutans} to read a count table and its index files. You can check
if two factor levels are correctly assigned using \Rfunction{pData}. Refer to
the help documentation of the function in R.
<<eval=FALSE>>=
help(readSmutans)
@
<<readSmutans>>=
smutansGenes <- 
  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
               indexFile =file.path(inDir, "count-ua159.txt.index"),
               condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
               firstFactor = c("UA159GLU", "UA159GAL"),
               firstFactorLabel = c("ua159", "tw1"),
               secondFactor = c("UA159GLU", "TW1GLU"),
               secondFactorLabel = c("glucose","galactose"),
               name="S. mutans UA159, TW1, Glucose, and Galactose")
@

The following functions create R data objects. Users are not recommended to use
them because they have been already created and saved together in the package.
They will be replaced by functions of \Rpackage{smutans}. They are documented
for the sake of reminders.
<<samplePreparation,eval=FALSE>>=
# not run
# but create a R scirpt file with the following scripts, and
# run it, i.e., Rscript run.R
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData34( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
smutans.prepareTranscript( )
q("no")
@
Now we look at the detail of the loaded count table. There are 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA-seq samples in \Robject{smutansGenes}, which
contains the counts and also metadata about the samples:
<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
@
<<look_into_smutansGenes>>=
pData( smutansGenes )
@
The samples differ by experimental condition (``glucose'' or ``galactose'') and
by library type (``ua159'' or ``tw1'').  Use the following functions to get
helps such as opening the vignette and manual of \Rpackage{smutans}. 
<<eval=FALSE>>=
help.start()
vignette(package="smutans")
openVignette("smutans")
@

%--------------------------------------------------
\section{DE analysis using UA159 and TW1 samples with glucose and galactose}
%--------------------------------------------------

We compared four pairs of experiments using the \Robject{smutansGenes}: two comparisons
between ``glucose'' and ``galactose'' given either type ``ua159'' or ``tw1'',
and another two comparisons between types ``ua159'' and ``tw1'' given either
condition ``glucose'' or ``galactose.'' Before comparing these we clustered
samples using read counts. Figure \ref{figClust} shows the cluster of the
\Sexpr{length(colnames(counts(smutansGenes)))} samples. Refer to
\Rpackage{DESeq} vignette as to how the clustering is conducted.  The clustering
was useful in ascertaining that the nutrient effect was stronger than the mutant
effect: two experimental conditions of ``glucose'' and ``galactose''
distinguished the samples. Within each condition, two different types of
``ua159'' and ``tw1'' separated them.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
as calculated from the variance-stabilizing transformation of the
count data.}
\label{figClust}
\end{figure}
<<figClust,fig=TRUE>>=
allSample <- newSmutans( smutansGenes )
smutans.de2Clust( allSample )
@

%------------------------------------------------------------
\subsection{Inference using single-factor data} 
%------------------------------------------------------------

Let us compare ``glucose'' and ``galactose'' conditions given ``ua159'' type. 
\Robject{smutansGenes}
contains short reads counts based on bwa alignment on UA159 genome.  The title
argument would be useful in distinguishing different objects of \Rclass{Smutans}
as we have more R objects of the same type. A subsequent call to
\Rfunction{smutans.de2} uses the two-factor data set to test differential
expression for the annotated genes by estimating
effective library sizes, and dispersions using functions provided by
\Rpackage{DESeq}.
<<ua159only>>=
ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
ua159only <- smutans.de2( ua159only, type="ua159", 
                          condA="glucose", condB="galactose" )
head(smutans.de2List( ua159only ))
@
Descriptions of the columns are in order.
Each gene has its mean expression level (at the base scale) as a joint estimate
from both conditions, and estimated separately for each condition, the fold
change from the first to the second condition, i.e.,
$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2) of the fold change,
and the $p$ value for the statistical significance of this change. The
\texttt{padj} column contains the $p$ values, adjusted for multiple testing with
the Benjamini-Hochberg procedure (see the \Rfunction{p.adjust}),
which controls false discovery rate (FDR). Later, we print the gene list in a
CSV file using option file of \Rfunction{smutans.de2List}. Windows users should
be able to open CSV files with a spreadsheet program such as EXCEL. Users who
are more comfortable with the EXCEL program than R could post-process CSV files
for sorting and finding differentially expressed genes using different q-value
cutoff values.
<<eval=TRUE>>=
smutans.de2List( ua159only, "ua159only.csv" )
@

By fixing ``tw1'' type similarly we compared ``glucose'' and ``galactose''
conditions.
<<tw1onlySamples>>=
tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
tw1only <- smutans.de2( tw1only, type="tw1", 
                        condA="glucose", condB="galactose" )
@
We can print the result in a csv file.
<<eval=TRUE>>=
# not run
smutans.de2List( tw1only, "tw1only.csv" )
@
We compared ``glucose'' and ``galactose'' with either ``ua159'' or ``tw1''
fixed. We computed numbers of differentially expressed genes that were
determined by both or either one of them.
<<>>=
tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
addmargins( tabUA159onlyTW1only )
@
The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
genes were found in both analyses. The analysis using TW1-only sample found
\Sexpr{tabUA159onlyTW1only[1,2]} among genes UA159-only sample could not
find. The analysis using UA159-only sample found
\Sexpr{tabUA159onlyTW1only[2,1]} genes among ones TW1-only sample could not
find. UA159-only sample detected more differentially expressed genes.

We now compare sample types on differentially expressed genes. Create two
\Rclass{Smutans} objects: \Robject{glucoseOnly} and \Robject{galactoseOnly}, and
compare them in detecting differentially expressed genes:
<<glucoseOnlySamples>>=
glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
                              condA="ua159", condB="tw1" )
tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
                                                      galactoseOnly )
addmargins( tabGlucoseOnlyGalactoseOnly )
@
Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
rate of 10\%. The analysis using galactose-only sample found
\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} of the genes that were negative in
glucose-only sample.  The analysis using glucose-only
sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} of the genes that were
negative in galactose-only sample. 
Table \ref{tab:de} summarizes the number of differentially expressed genes that
were found using single-factor analyses at FDR of 10\%. We can print them in csv
files using \Rfunction{smutans.de2List}.
<<eval=TRUE>>=
# not run
smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
smutans.de2List( galactoseOnly, "galactoseOnly.csv" )
@

%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Numbers of differentially expressed genes in different analyses}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{ccccl}
\hline 
Sample A & Sample B & Number of Genes & Result File \\
\hline 
UA159 Glucose & UA159 Galactose & \Sexpr{sum(ua159only@padj < .1,na.rm=T)} & ua159only.csv \\
TW1 Glucose & TW1 Galactose & \Sexpr{sum(tw1only@padj < .1,na.rm=T)} & tw1only.csv \\
UA159 Glucose & TW1 Glucose & \Sexpr{sum(glucoseOnly@padj < .1,na.rm=T)} & glucoseOnly.csv \\
UA159 Galactose & TW1 Galactose & \Sexpr{sum(galactoseOnly@padj < .1,na.rm=T)} & galactoseOnly.csv \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:de}
\end{table}
%-----------------------------------------------------------------------------

%------------------------------------------------------------
\subsection{Inference using all of the samples} 
%------------------------------------------------------------

Let us determine genes differentially expressed in different conditions of
``glucose'' and ``galactose'' using both types of ``ua159'' and ``tw1'': a
two-factor analysis. Start
with the two-factor analysis by creating a new \Rclass{Smutans} object, called
\Robject{allSample}. Although we already used \Robject{allSample} when
clustering the samples, we recreate it for instructive purposes.  The
\Rfunction{smutans.de2} can use a two-factor data set to perform two-factor
analysis: we are here interested in condition effect. 
<<allSamples,results=hide>>=
allSample <- newSmutans( smutansGenes, 
                         title="All Samples for Comparing Conditions" )
allSample <- smutans.de2( allSample,
                          condA="glucose", condB="galactose" )
tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
                                                   allSample )
@
<<>>=
addmargins( tabUA159onlyAndAllSample )
@
Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. 
Using all of the samples detected more differentially expressed genes than did
ua159-only sample.
The output from \Robject{allSample} is almost the same except that there are
convergence column.
<<>>=
head( smutans.de2List( allSample ) )
@

We can show that taking the library type into account was important to have
good detection power by doing the analysis againt using the standard workflow,
as outlined earlier, and without informing \Rpackage{DESeq} of the library
types. This anlaysis is not necessary. However, it clearly shows the advantage of the
two-factor analysis.
<<allSampleWithoutType>>=
allSampleWithoutType <- newSmutans( smutansGenes, "All Sample without Type" )
allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
                                                condA="glucose", 
                                                condB="galactose" )
tabWithoutWithAllSample <- smutans.compareResult( allSampleWithoutType, 
                                                  allSample )
addmargins( tabWithoutWithAllSample )
@
Using all of the sample that considers types found more differentially expressed
genes than did one ignoring the types.

We now proceed to compare the two types by considering all of the samples, and
compare the result with the single-factor analysis.
<<allSamplesType,results=hide>>=
allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
allSampleType <- smutans.de2( allSampleType, compareCondition="no",
                              condA="ua159", condB="tw1" )
tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
                                                  allSampleType )
tabGalactoseOnlyAllSample <- smutans.compareResult( galactoseOnly, 
                                                    allSampleType )
@                                                    
<<>>=
addmargins( tabGlucoseOnlyAllSample )
addmargins( tabGalactoseOnlyAllSample )
@
Of the total \Sexpr{sum(tabGlucoseOnlyAllSample)} genes,
\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
of the genes that glucose-only sample did not find. 
The analysis using glucose-only sample
found \Sexpr{tabGlucoseOnlyAllSample[2,1]} the genes that all samples did not
find.  In the comparison using galactose-only sample,
of the \Sexpr{sum(tabGalactoseOnlyAllSample)} genes,
\Sexpr{tabGalactoseOnlyAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabGalactoseOnlyAllSample[1,2]}
the genes galactose-only sample did not find. 
The analysis using galactose-only sample
found \Sexpr{tabGalactoseOnlyAllSample[2,1]} the genes that all samples did not
find.

One can print the sults in CSV files.
<<eval=TRUE>>=
smutans.de2List( allSample, "allSample.csv" )
smutans.de2List( allSampleType, "allSampleType.csv" )
@

%------------------------------------------------------------
\subsection{Functional category association analyses} 
%------------------------------------------------------------

We find some functional categories enriched with differentially expressed genes
using \Rpackage{goseq} \cite{Young2010} (table \ref{tab:goseq}).
To find functional categories enriched with differentially expressed genes we
first find a list of differentially expressed genes. A set of genes are assigned
to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
UA159 include gene lengths: \Robject{smutans.feature.genes}.
Functional categories with numbers of genes assigned are in
\Robject{smutans.cat.desc}.
<<goseqUA159only,eval=FALSE>>=
data( "smutans.feature.genes" )
data( "smutans.go.genes" )
data( "smutans.cat.desc" )
smutans.de2Goseq ( ua159only, qval=1e-2, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<goseqUA159only,echo=FALSE,results=tex>>=
data( "smutans.feature.genes" )
data( "smutans.go.genes" )
data( "smutans.cat.desc" )
smutans.de2Goseq ( ua159only, qval=1e-2, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:goseq}
\end{table}
%-----------------------------------------------------------------------------

%--------------------------------------------------
\section{DE analysis using 835NP, 835P, and UA159 no CSP}
%--------------------------------------------------

Sang-Joon Ahn is leading this experiment. Let us create the input data variable.
We compare three comparisons using three conditions. We do the cluster,
determine differentially expressed genes, and print those genes.  One of the
three conditions, 835P, is distinguished from the others.  We found no
functional categories enriched with differentially expressed genes in the
pairwise comparisons of 835NP, 835P, and UA159 no CSP.  Let's make a list of
differentially expressed genes.  
<<readSmutans>>=
sm835Genes <- 
  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
                indexFile =file.path(inDir, "count-ua159.txt.index"),
                condition = c("835NP", "835P", "UA159noCSP"),
                firstFactor = c("835NP", "835P", "UA159noCSP"),
                firstFactorLabel = c("ua159"),
                secondFactor = c(),
                secondFactorLabel = c("835np","835p", "ua159nocsp"),
                name="S. mutans 835NP, 835P, UA159 no CSP")
@

<<>>=
head( counts(sm835Genes), n=1 )
pData( sm835Genes )
@

<<figClustNpAndP,fig=TRUE>>=
npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
smutans.de2Clust( npAndP )
@
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength.}
\label{figClustNpAndP}
\end{figure}
We first cluster the \Sexpr{length(colnames(counts(sm835Genes)))} samples
(fig.\ \ref{figClustNpAndP}).
